EAST: {\underline E}xpression Fragment {\underline A}ssembly from {\underline S}panning {\underline T}rees

### 2011-01-10 peace/papers/EAST/east.tex
First version of the EAST manuscript (FINALLY)

\author{Yuan Zhang$^1$
  \email{Yuan Zhang - zhangy9@muohio.edu}
  \and
  James C Moler$^1$
  \email{James Moler - molerjc@muohio.edu}
  \and
  Dhananjai M Rao$^1$
  \email{Dhananjai Rao - raodm@muohio.edu}
  \and
  Mufit Ozden$^1$
  \email{Mufit Ozden - ozdenm@muohio.edu}
  \and
  Chun Liang\correspondingauthor$^{1,2}$
  \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  and
  John E. Karro\correspondingauthor$^{1,3}$
  \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Introduction}
The {\it de novo} assembly of RNA transcripts is an important and
computationally challanging problem, made more difficult by the advent
of Next Generation Sequencing technologies such as 454 or Illumina
[CITE].  Given the completion of large-scale seqencing run of multiple
mature transcripts of a given organism, the invetigator is left with a
large set of sequence fragments, but no ordering or even transcript
association information; these fragments must be clustered by
transcript then assembled before the sequencing output can be truely
useful.  If the genome has not been sequenced, this must be done {\it
  de novo} -- using only the sequence information extrapolated from
the sequences themselves.  Even when the genomic sequence is known,
there are a number of advanatages to ignoring the known sequence
information and starting from scratch.  Using an existing squence as a
template for sequencing in population studies has the potential to
introduce bias.  And when looking specifically for genes, a known
sequence may be of limited use: the target transcripts may be
undiscovered, and the splicing of introns before sequencing can complicate
the approach of mapping fragemtns to the genomic sequence.  In short,
there is significant value in tools that can reassemble fragement data
into the orginial assembly without requiring access to genomic
information.

The problem of {\it de novo} transcript assembly has been addressed a number
of times in the literature, with initial approaches consentrating on
the longer reads produced by traditional Sanger Sequencing techniques,
and newer approaches looking at produces of Next Generation Sequencing
(NGS) technologies.  A list of the tools most successfully addressing
these problems include \capthree, \tgicl, \velvet, and \mira [CITE],
each excelling within their own context.  \capthree\/ and \tgicl\/ use
alignment-based strategies that produce high quality results when
applied to the output of Sanger sequencing technologies.  \velvet,
based on the modeling of the problem with a {\it de bruijn} graphs,
was designed to handle short read sequences.  [SOMETHING ABOUT MIRA --
INCLUDING ABIITY TO HANDLE HYBRID DATA.]

In this paper we present \east, a tool for the assembly of both
Sanger, NGS and hybrid data with the capacity to incorperate quality
scores when available.  \east\/ is based on a novel minimum spanning
tree (MST) apraoch to compute assembly.  The tool works in tandem with
the \peace\/ clustering tool [CITE] -- allowing \peace\/ to cluster
fragements by transcript and produce the initial MSTs that will serve
as the basis for the \east\/ algorithm.  In the following we present a
comparison of \east\/ against those of \capthree, \tgicl, \velvet, and
\mira on both Sanger and NGS sequences, showing that [???]

@@

In assessing result quality, we must view the \peace\/ clustering tool
\cite{Rao10} and \east\/ as a combined process.  While most assembly
tools integrate these two jobs, \east\/ assumes clustering has been
preformed and requires the \peace\/ MST output.  Hence for purposes of
the following section, we consider a single tool \peast, and
assess the quality of these results.

In order to assess the quality of the \east\/ (\peast\/) results, we compared
against four prominent assembly tools: \capthree, \tgicl,
\mira, and \velvet [CITE].  Tools were tested on data from three
sequencing technologies (Sanger, 454 and Illumina), using both
simulated output obtained from the \estsim\/ and \metasim\/ tools
[CITE] and benchmark sets obtained from GenBank [CITE].

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized A-score}, based
on the {\it assembly score} (A-score) used in {\it Liang et al.}
\cite{Liang00}.  For an assembly of some originial sequence, the
A-score is defined as:
$$\mbox{($2\times$sequence length) - ($5\times$ number of
mismatches) - ($15 \times$ number of indels)}$$
with a ``perfect'' score equal to twice the sequence length.  Our
normalized A-score modifies this definition as:
$$\frac{\mbox{($2\times$assembly length) - ($5\times$ number of mismatches)
  - ($15\times$ number of indels)}}{\mbox{($2\times$ sequence length)}}$$
with a perfect score equal to 1.  This normalization allows for the
comparison of assembly quality between sequences of differenct
lengths. Further, using this score we can more accurately quantify the
quality of a partial reconstrction: if we are able to reconstruct only
a portion of a fragment, the normalized A-score will take full
consideration of the quality of the reconstruction, but then bound the
score by the relative size of the reconstruction.  (e.g. Perfectly
reconstructiong one fourth of the original sequence will lead to an a
score of $0.25$.)

In addition to A-score...

\subsection*{Simulation Results}
Using simulated fragement sets allows us to both compare assemblies to
a known correct solution and control parameters in their model
(e.g. base-call error rate) to test the effect of variation.  All sets
were generated from a set of 100 zebra fish transcripts obtained from
{\it Hazelhurst et al.}  \cite{Hazelhurst08}.  For simulated Sanger
sequences we used the \estsim\/ tool, using its default model of EST
generation and error rate when not otherwise specified.  For simulated
454 and Illumina sets we used the \metasim\/ tool, again using their
default model to simulate the errors associated with those
technologies.

\noindent {\bf Sanger Sequencing:} We start by looking at the assembly
quality of each tool, when applied to simulated Sanger Sequences, as a
function of error rate for a range of coverage depths.  For each
parameter combination we applied each tool to 30 independently
randomly generated simulated sets, derived from 15 zebra fish genes,
reporting the average normalized A-score.  In
Figure~\ref{sangerAscore} we look at normaized A-score as a function
of error rate, fixing coverage at 50, 30 and 10 and varying base-read
error rate from 0\% to 6\% [WHAT IS INDEL RATE?] for each of the five
tools being tested.  While the three illustrated tools preform well
when applied error free (i.e. impossible) sequencing technoloy, we
find \east\/ to be considerably more robust to base-call errors than the
other tools.  Taking the mid-range coverage value of 30 as an example,
at a 1\% error rate we see \east, \capthree, and \tgicl\/ all achieving an
essentially perfect normalize A-score, while \mira\/ ranges around a 0.5
and \velvet\/ does quite poorly.  At a 3\% error rate \east\/ maintains its
almost perfect normalized A-score, showing a 2\% improvement of \tgicl,
a 21\% improvlement over \capthree, and a 133\% improvement over \velvet.
At a 4\% error rate \east\/ still maintains its near-perfect score
(showing a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
respecitvally), and even at a 6\% error rate is still achieving an
normalized A-score of 0.95 (a 110\% improvement over the next closest
tool, \tgicl\/).

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger.png}}
\label{sangerAscore}
\caption{Normalized a-score as a function of base error rate for
  simmulated sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.}
\end{figure}

Another useful metric is the number of contigs and number of singleton
fragments in the final result.  In Figure~\ref{contigs.sanger} is a
plot of the average number of contigs per gene seqence (left column), ad
the number of singletons (right column).  For the first we observe that all tools have
comperable results for lower error rates, but only \east\/ is able to
maintain the low score with increasing error rates.  In minimizing the
number of singletons, we find that \east\/ does a better job than
\capthree\/ for error rates larger than $2\%$, while TGICL does a
better job that \east\/ for error rates larger than $4\%$ to $5\%$.

\begin{figure}[htb]
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numContigs_sanger.png}
\end{minipage}
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numSingle_sanger.png}
\end{minipage}
\label{contigs.sanger}
\caption{Number of contigs and singles as a function of base error
  rate for simmulated sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.  (Number of singles for
  \velvet\/ and \mira\/ not shown.)}
\end{figure}

[ADD SOMETHING ABOUT COVERAGE?  The situation is strange.]

\noindent {\bf Next Generation Sequences:}

@@

@@

The presence of an edge in the tree is a heuristic indication that the
incident sequences overlap; were there not sufficient sequence
similarity to indicate an overlap, the edge would have been assigned a
high $d^2$ value and would have been removed by \peace\/.  To
determine how the nodes overlap, we calculate a modified $d^2$ score
as follows.  The standard calculation of $d^2$, as define by {\it Hide
  et al.} \cite{Hide94}, requires looking at every pair of length $w$
sub-strings (windows) from the two sequences, calculating a distance
between these two windows based on the number of shared words
(6-mers), and taking the score of the lowest-scoring window pairs as
the $d^2$ distance.  However, when looking for overlap of fragments
$s_1$ and $s_2$, we are looking only for end-overlap.  That is, we
wish to determine of the left end of $s_1$ overlaps the right end of
$s_2$, or the reverse (recalling that we know them to be on the same
strand).  Hence we need only consider the left-most and right-most
windows of $s_1$ against each window of $s_2$, reducing the number of
window comparisons from quadratic to linear in the size of $s_2$.

In Figure~\ref{fig:overlap} we see illustrate the use of modified $d^2$,
using it to take two sequences (a), find the window on $s_2$ most
resembling the right-most window of $s1$ (red section of (b)),
extending to the end of $s_2$ (green section of (c)) and preforming a
{\it global} alignment on the overlapping portion.  We then compute the
{\it overlap distance} ($1 - s/l$, where $s$ is the alignment score and
$l$ is the length of the overlap segment), a value that is inversely
proportional to the probability that this is a legitimate overlap.

@@

We note that both \peace\/ and \east\/ make use of an {\it adaptive
  $d^2$} strategy to handle hybrid data sets.  The $d^2$ measure (and
our modification) is flexible enough to handle a range of sequence
sizes, requiring only a modification of the parameters (i.e. window
size and threshold values).  However, it runs into problems when
comparing sequences of significantly different sizes.  Comparing a 62
base Illumina read against a 1000 base Sanger read requires the use of
parameters appropriate to the shorter fragment -- which are
considerably less accurate when applied to the longer read.  Our
solution is to partition fragments by sizes (e.g. into groups of
small, medium and large fragments), and compute distances only
between segments in the name or neighboring groups.  By skipping the
comparisons between sequences of significantly different sizes we
avoid the pitfalls of such comparisons.

@@

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )

@@

\section*{Figures}
  \subsection*{Figure 1 - Sample figure title}
      A short description of the figure content
      should go here.

  \subsection*{Figure 2 - Sample figure title}
      Figure legend text.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Sample table title}
    Here is an example of a \emph{small} table in \LaTeX\ using  
    \verb|\tabular{...}|. This is where the description of the table 
    should go. \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        A1 & B2  & C3 \\ \hline
        A2 & ... & .. \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

@@

% LocalWords:  Ozden Chun Liang Karro Botanty publ MSTs et al Hazelhurst Prim's
% LocalWords:  mers Illumina contig bmc multi

>>>>

\author{Yuan Zhang$^1$
  \email{Yuan Zhang - zhangy9@muohio.edu}
  \and
  Dhananjai M Rao$^1$
  \email{Dhananjai Rao - raodm@muohio.edu}
  \and
  Mufit Ozden$^1$
  \email{Mufit Ozden - ozdenm@muohio.edu}
  \and
  James C Moler$^1$
  \email{James Moler - molerjc@muohio.edu}
  \and
  Chun Liang\correspondingauthor$^{1,2}$
  \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  and
  John E. Karro\correspondingauthor$^{1,3}$
  \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Introduction}
The {\it de novo} assembly of RNA transcripts is an important and
computationally challenging problem, made more difficult by the advent
of Next Generation Sequencing technologies such as 454 or Illumina
\cite{Nagaraj07,Rao10}.  Given the completion of a large-scale sequencing run of
multiple mature transcripts of a given organism, the investigator is
left with a large set of sequence fragments, but no ordering or even
transcript association information; these fragments must be clustered
by transcript then assembled before the sequencing output can be truly
useful.  If the genome has not been sequenced, this must be done {\it
  de novo} -- using only the sequence information extrapolated from
the sequences themselves.  Even when the genomic sequence is known,
there are a number of advantages to ignoring the known sequence
information and starting from scratch.  Using an existing sequence as
a template for sequencing in population studies has the potential to
introduce bias.  And when looking specifically for genes, a known
sequence may be of limited use: the target transcripts may be
undiscovered, and the splicing of introns before sequencing can
complicate the approach of mapping fragments to the genomic sequence.
In short, there is significant value in tools that can reassemble
fragment data into the original assembly without requiring access to
genomic information.

The problem of {\it de novo} transcript assembly has been addressed a
number of times in the literature, with initial approaches
concentrating on the longer reads produced by traditional Sanger
Sequencing techniques, and newer approaches looking at produces of
Next Generation Sequencing (NGS) technologies.  A list of the tools
most successfully addressing these problems include \capthree, \tgicl,
\velvet, and \mira, each excelling within their own context
\cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  \capthree\/ and
\tgicl\/ use alignment-based strategies that produce high quality
results when applied to the output of Sanger sequencing technologies.
\velvet, based on the modeling of the problem with a {\it de bruijn}
graphs, was designed to handle short read sequences.  \mira\/ is a
multi-pass DNA assembler based on an overlap-layout-consensus
strategy.  [STILL NEED TO SAY SOMETHING ABOUT MIRA AND HYBRID SETS.]

In this paper we present \east, a tool for the assembly of both
Sanger, NGS and hybrid data with the capacity to incorporate quality
scores when available.  \east\/ is based on a novel minimum spanning
tree (MST) approach to compute assembly.  The tool works in tandem with
the \peace\/ clustering tool \cite{Rao10}: \peace\/ clusters
fragments by transcript and produces the initial MSTs that will serve
as the basis for the \east\/ algorithm.  In the following we present a
comparison of \east\/ against those of \capthree, \tgicl, \velvet, and
\mira on both Sanger and NGS sequences, showing that \peast\/
generates the best quality results in the fastest runtime for Sanger
sequences, and that while certain tools out-preform \peast\/ under specific
conditions, it is the only tool able to produce constant quality results
for Sanger, 454, Illumina, {\it and} hybrid data sets.

@@

In assessing result quality, we must view the \peace\/ clustering tool
\cite{Rao10} and \east\/ as a combined process.  While most assembly
tools integrate these two jobs, \east\/ assumes clustering has been
preformed and requires the \peace\/ MST output.  While other cluster
tools could be substituted in to preform the actual clustering
(e.g. \wcd or \easycluster \cite{Hazelhurst08,Picardi09}), only
\peace\/ provides the requires MST output.  Hence for purposes of the
following section, we consider a single tool \peast, and assess the
quality of these results.

In order to assess the quality of the (\peace\/ +) \east\/ results, we compared
against four prominent assembly tools: \capthree, \tgicl,
\mira, and \velvet\/ \cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  Tools were tested on data from three
sequencing technologies (Sanger, 454 and Illumina), using both
simulated output obtained from the \estsim\/ and \metasim\/ tools
\cite{Hazelhurst03,Richter08} and benchmark sets obtained from GenBank \cite{Benson10}.

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized a-score}, based
on the {\it assembly score} (a-score) used in {\it Liang et al.}
\cite{Liang00}.  For an assembly of some original sequence, the
a-score is defined as:
$$\mbox{($2\times$sequence length) - ($5\times$number of
mismatches) - ($15 \times$number of indels)}$$
with a ``perfect'' score equal to twice the sequence length.  Our
normalized a-score modifies this definition as:
$$\frac{\mbox{($2\times$assembly length) - ($5\times$number of mismatches)
    - ($15\times$number of indels)}}{\mbox{($2\times$sequence
    length)}}$$ with a perfect score equal to 1.  This normalization
allows for the comparison of assembly quality between sequences of
different lengths. Further, using this score we can more accurately
quantify the quality of a partial reconstruction: if we are able to
reconstruct only a portion of a fragment, the normalized a-score will
take full consideration of the quality of the reconstruction, but then
bound the score by the relative size of the reconstruction.
(e.g. Perfectly reconstruction one fourth of the original sequence
will lead to an a score of $0.25$.)  In addition to a-score we also
look at: the number of contigs produced by a tool (which, in the
perfect solution, will be exactly equal to the number of transcripts
represented by the data set), normalized by the number of transcripts;
the number of singleton ESTs in the solution (i.e. ESTs unassigned to
any contig and thus, most cases, effectively thrown away), and the runtime of the tool.

\subsection*{Simulation Results}
Using simulated fragment sets allows us to both compare assemblies to
a known correct solution and control parameters in their model
(e.g. base-call error rate) to test the effect of variation.  All sets
were generated from a set of 100 zebra fish transcripts obtained from
{\it Hazelhurst et al.}  \cite{Hazelhurst08}.  For simulated Sanger
sequences we used the \estsim\/ tool, using its default model of EST
generation when not otherwise specified.  For simulated
454 and Illumina sets we used the \metasim\/ tool \cite{Richter08}, again using their
default model to simulate the errors associated with those
technologies.

\noindent {\bf Sanger Sequencing:} We first compare the assembly
quality of the tools on Sanger Sequencing, using \east, \capthree, and
\tgicl\/ (the tools designed to work with Sanger sequences), looking
at the a-score as a function of error rate for a range of coverage
depths.  For each parameter combination we applied each tool to 30
independently randomly generated simulated sets, derived from 15 zebra
fish genes, reporting the average normalized a-score.  In
Figure~\ref{sangerAscore} we look at normalized a-score as a function
of error rate, fixing coverage at 50, 30 and 10 and varying base-read
error rate from 0\% to 6\% for each of the three tools designed for
Sanger Sequencing.  While all three tools preform well when applied to
error free (i.e. impossible) sequencing technology, we find \east\/ to
be considerably more robust to base-call errors than the other tools.
Taking the mid-range coverage value of 30 as an example, at a 1\%
error rate we see \east, \capthree, and \tgicl\/ all achieving an
essentially perfect normalize a-score .  At a 3\% error rate \east\/
maintains its almost perfect normalized a-score, showing a 2\%
improvement over \tgicl\/ and a 21\% improvement over \capthree.  At a
4\% error rate \east\/ still maintains its near-perfect score (showing
a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
respectively), and even at a 6\% error rate is still achieving an
normalized a-score of 0.95 (a 110\% improvement over the next closest
tool, \tgicl\/), meaning that at most 5\% of the bases were left out of
their contigs or were incorrectly reconstructed.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger.pdf}}
\caption{Average normalized a-score as a function of base error rate for
  simulated Sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.}
\label{sangerAscore}
\end{figure}

Another useful metric is the number of contigs and number of singleton
fragments in the final result.  In Figure~\ref{contigsSanger} is a
plot of the average number of contigs per gene sequence (left column), and
the number of singletons (right column).  For the first we observe that all tools have
comparable results for lower error rates, but only \east\/ is able to
maintain the low score with increasing error rates -- generating an
perfect score in almost every case.  In minimizing the
number of singletons, we find that \east\/ does a better job than
\capthree\/ for error rates larger than $2\%$, while \tgicl does a
better job that \east\/ for error rates larger than $4\%$ to $5\%$.

\begin{figure}[htb]
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numContigs_sanger.pdf}
\end{minipage}
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numSingle_sanger.pdf}
\end{minipage}
\caption{Number of contigs (normalized to the number of transcripts)
  and number of singletons as a function of base call error
  rate for simulated Sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.  (Number of singles for
  \velvet\/ and \mira\/ not shown.)}
\label{contigsSanger}
\end{figure}

\noindent {\bf Next Generation Sequences:} In
Figure~\ref{nextgenAscore} we compare the normalized a-score for
simulated 454 and Illumina outputs (with sequence lengths averaging
250 and 62 bp respectively).  For these tests we varied coverage
levels but otherwise took the default Metasim model.  On 454 data we
see significant improvement in results quality in \east\/ for all
levels of coverage.  On Illumina data we see that \east\/ is
competitive with \velvet\/ at higher coverages, though is unable to
match that tool when coverage is sparse.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_nxtgen.pdf}}
\caption{Comparative normalized a-scores for 454 and Illumina sequence
  output, using the \metasim default model at different levels
  of coverage.  \east=blue, \capthree=Black, \tgicl=green,
  \velvet=red, \mira=purple.  }
\label{nextgenAscore}
\end{figure}

\noindent {\bf Hybrid:} In Figure~\ref{hybridAscore} we look at the
quality of results derived from applying each tool to a hybrid data
set of Sanger (40\%), 454 (40\%) and Illumina (20\%) sequences, again
showing normalized a-score as a function of base call error rate in
the Sanger portion.  Here we observe \east matching or outperforming
\capthree\/ and \mira\/ at moderate to high coverages.  At a coverage
level of 50, \east\/ maintains a normalized a-score hovering around
0.85 at all error levels (showing a 54\% improvement over \mira\/ at a
3\% base-call error rate).  \east\/ actually improves its performance
at a coverage level of 30 (hovering around a normalized a-score of
0.95, showing a 142\% improvement of Mira at the 3\% base-call error
rate), an oddity we attribute to the decreased absolute number of NGS
sequences with the smaller coverage.  At a coverage level of 10
\east's normalized a-score is around 0.37 -- doing considerably worse
that \mira\/ at low error rates (e.g. 0.94 v. 0.40 for error-free
bases), but showing a 52\% improvement over \mira\/ at the 3\% error
rate mark and in general showing much more robustness to error.
Remarkably, it is only in this situation that \capthree\/ is
competitive, with that tool achieving a normalized a-score of 0.62
(v. \east's 0.37) for the 3\% base call error rate, but loosing out to
\east at greater than a 4.5\% base call error rate -- a result that we
attribute to \east's problems with the 62 bp Illumina sequences (see below).

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_hybrid.pdf}}
\caption{Comparative normalized a-scores a hybrid set of Sanger, 454
  and Illumina sequence output, varying the base call error rate for
  Sanger sequences and using the MetaSim default model for the 454 and
  Illumina sequences.  Each data set tested was composed of 40\%
  Sanger Sequences, 40\% 454 Sequences, and 20\% Illumina sequences.
  \east=blue, \capthree=Black, \tgicl=green,
  \velvet=red,\mira=purple.}
\label{hybridAscore}
\end{figure}

\noindent {\bf Tool Runtime:} As \east\/ assembly requires the
\peace\/ cluster (as opposed to the competing tools that do both
jobs), for \east\/ runtime we actually include the \peace\/
runtime. In Figures~\ref{runtime.fixed} we see the average runtime for
each of the experiments reflected in Figures
Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30,
while in Figure~\ref{runtime.coverage} we look at the average runtime
for each tool when applied to simulated data sets generated from a
single zebra fish gene at varying levels of coverage -- effectively
plotting the runtime as a function of fragment set size.  For Sanger
sequences, we see marked improvement of (\peace+)\east\/ over all
other tools for almost all coverage levels: for the smallest set
(coverage = 10) \east\/ has 1.9 speedup over \capthree\/ and a 3.9
speedup over \tgicl, while at a more moderate size (coverage=50) we
see speedups of 4.7 and 8.4.  This returns back down to 3.0 and 5.2 at the
largest set tested (coverage = 100), as can be seen in the bump around
a coverage of 85 in Figure~\ref{runtime.coverage}.  We are unable to
superficially explain this bump, but have noticed the same effect in
data sets derived from other genes as well.

\east\/ is less competitive on shorter sequence (an effect of the
$d^2$ bias of our algorithm, as discussed below), reducing its runtime
improvement for the 250 bp 454 sequences, severely reducing it for the
62 bp Illumina sequences, and also affecting the hybrid data sets (also
due to the presence of the Illumina sequence).  We argue that the high
quality of our results still makes \east\/ an appealing option for 454
and hybrid data (while acknowledging that cannot complete with
\velvet\/ on pure Illumina data), and also observe that while it has
not been implemented, the basic \east\. algorithm should easily admit
paralilization -- thus increasing the practical limit on data set size.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
\caption{Average runtime per trial for when generating data for
  Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30).
  Note that the exceptionally high speed of \velvet\/ on Illumina sequences make the
  red (center) bars in that plot essentially indistinguishable.
  (\peace+)\east=blue, \capthree=Black, \tgicl=green,
  \velvet=red,\mira=purple.}
\label{runtime.fixed}
\end{figure}

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_bycoverage_all.pdf}}
\caption{The average runtime when applying each tool to a simulated
  data set derived from a specific zebrafish transcript of length
  14,625 bp.  As we decrease coverage the number of fragments drops,
  hence illustrating the relationship between tool runtime and
  fragment set size.  Each data point represents the average runtime
  from the application of the tool to at least 30 randomly generated
  simulated sets coming from the transcript at the given level of coverage.
  (\peace + )\east=blue, \capthree=Black, \tgicl=green,
  \velvet=red,\mira=purple.}
\label{runtime.coverage}
\end{figure}

@@

In this manuscript we have described the \east\/ algorithm and run a
comparison of the \peast combination against the best preforming tools
in the literature.  When compared against the fine benchmark tools on
normalized a-score and runtime metrics, the \peast\/ combination of
tools prove to be both the best tool for Sanger sequencing, the
best tool overall, and by far the most robust against sequencing
error.  While certain tools may do better than it in specific circumstances,
there is no tool that does better than it on multiple sequencing
technology platforms.  While its speed is slower when applied to short
sequence fragments, the quality of its results more then compensate.

As observed in the results section, when applying \east, \capthree,
and \tgicl\/ to Sanger sequenced transcripts with normal to high error
rates, \east\/ consistently out preforms all other tools in terms of
the normalized a-score metric when applied to transcript fragments
resulting from Sanger sequencing (Figures~\ref{sangerAscore},
\ref{runtime.fixed}, and \ref{runtime.coverage}).  We also find that
the number of contigs produced is almost exactly equal to the number
of transcripts represented in the set, regardless of error rate, while
the number of singleton ESTs remains low up to about a 5\% error rate
(Figures~\ref{contigsSanger}).  The high normalized a-score indicates
that \east\/ is both reconstructing larger portions of each transcript
and doing so more faithfully, while the low number of contigs means
that we are getting complete transcripts.  Finally, runtimes for
\east\/ are considerably faster than other tools for larger sets
(Figures~\ref{runtime.fixed} and \ref{runtime.coverage}, with
(\peace+)\east\/ running a minimum of two to three times as fast as
the other tools, with speedups over \capthree\/ (the faster tool)
ranging from 1.9 to 6.0, depending on the coverage level.  In short,
when used for Sanger Sequences \east\/ is a faster tool that produces
the high quality reconstructions versus comparing tools and is
significantly more robust to error than other tools.

When involving shorter sequences, the runtime of \east\/ takes a
significant hit.  As explained in our methods section, \east\/ is
dependent on the $d^2$ sequence distance function \cite{Hide94}, whose
parameters must be adjusted when applied to smaller sequences to
preserve quality, at the cost of runtime.  We describe below our {\it
  adoptive $d^2$} algorithm that allows us to target more specific
parameters on a case-by-case basis, but large numbers of short
sequences will still lead to slower runtimes.  In our tests our
simulated 454 sequences averaged 250 bp in length, while our Illumina
sequences were uniformly 62 bp, accounting for the slightly reduced
improvement of \east, as compared to other tool runtimes, in the 454 tests and
the significant reduction for the Illumina and Hybrid test sets
containing the short Illumina sequences.

Despite the slower runtime, it can still be argued that the \east\/ is
competitive, given the quality of its output, excepting when compared
to \velvet\/ on Illumina data.  For 454 sequences, no tool comes even
close to matching the \east\/ results.  \east is able to match
\velvet for high-coverage Illumina data, though loses out at lower
coverages (and is not even remotely competitive in terms of runtime).
Looking at hybrid sets, which the \mira\/ tool is specifically geared
to address, we again find \east\/ to produce significantly superior
results at moderate to high levels of coverage.  At low coverage
levels \east\/ does suffer (another effect of the short
Illumina sequences), but its robustness then leads to
better results when the Sanger sequence portion of the set is subject
to elevated error rates.

@@

The presence of an edge in the tree is a heuristic indication that the
incident sequences overlap; were there not sufficient sequence
similarity to indicate an overlap, the edge would have been assigned a
high $d^2$ value and would have been removed by \peace\/.  To
determine how the nodes overlap, we calculate a modified $d^2$ score
as follows.  The standard calculation of $d^2$, as defined by {\it Hide
  et al.} \cite{Hide94}, requires looking at every pair of length $w$
sub-strings (windows) from the two sequences, calculating a distance
between these two windows based on the number of shared words
(6-mers), and taking the score of the lowest-scoring window pairs as
the $d^2$ distance.  However, when looking for overlap of fragments
$s_1$ and $s_2$, we are looking only for end-overlap.  That is, we
wish to determine of the left end of $s_1$ overlaps the right end of
$s_2$, or the reverse (recalling that we know them to be on the same
strand).  Hence we need only consider the left-most and right-most
windows of $s_1$ against each window of $s_2$, reducing the number of
window comparisons from quadratic to linear in the size of $s_2$.

In Figure~\ref{fig:overlap} we illustrate the use of modified $d^2$,
using it to take two sequences (a), find the window on $s_2$ most
resembling the right-most window of $s1$ (red section of (b)),
extending to the end of $s_2$ (green section of (c)) and preforming a
{\it global} alignment on the overlapping portion.  We then compute the
{\it overlap distance} ($1 - s/l$, where $s$ is the alignment score and
$l$ is the length of the overlap segment), a value that is inversely
proportional to the probability that this is a legitimate overlap.

@@

We note that both \peace\/ and \east\/ make use of an {\it adaptive
  $d^2$} strategy to handle hybrid data sets.  The $d^2$ measure (and
our modification) is flexible enough to handle a range of sequence
sizes, requiring only a modification of the parameters (i.e. window
size and threshold values).  However, it runs into problems when
comparing sequences of significantly different sizes.  Comparing a 62
base Illumina read against a 1000 base Sanger read requires the use of
parameters appropriate to the shorter fragment -- which are
considerably less accurate when applied to the longer read.  Our
solution is to partition fragments by sizes (e.g. into groups of
small, medium and large fragments), and compute distances only
between segments in the name or neighboring groups.  By skipping the
comparisons between sequences of significantly different sizes we
avoid the pitfalls of such comparisons, while still picking up the
connections through transitivity.

@@

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{all} }     % Bibliography file (usually '*.bib' )

@@

% \section*{Figures}
%   \subsection*{Figure 1 - Sample figure title}
%       A short description of the figure content
%       should go here.

%   \subsection*{Figure 2 - Sample figure title}
%       Figure legend text.

@@

%% Use of \listoftables is discouraged.
%%
% \section*{Tables}
%   \subsection*{Table 1 - Sample table title}
%     Here is an example of a \emph{small} table in \LaTeX\ using  
%     \verb|\tabular{...}|. This is where the description of the table 
%     should go. \par \mbox{}
%     \par
%     \mbox{
%       \begin{tabular}{|c|c|c|}
%         \hline \multicolumn{3}{|c|}{My Table}\\ \hline
%         A1 & B2  & C3 \\ \hline
%         A2 & ... & .. \\ \hline
%         A3 & ..  & .  \\ \hline
%       \end{tabular}
%       }
%   \subsection*{Table 2 - Sample table title}
%     Large tables are attached as separate files but should
%     still be described here.

@@

% \section*{Additional Files}
%   \subsection*{Additional file 1 --- Sample additional file title}
%     Additional file descriptions text (including details of how to
%     view the file, if it is in a non-standard format or the file extension).  This might
%     refer to a multi-page table or a figure.

%   \subsection*{Additional file 2 --- Sample additional file title}
%     Additional file descriptions text.

@@

% LocalWords:  Ozden Chun Liang Karro Botanty publ MSTs et al Hazelhurst Prim's
% LocalWords:  mers Illumina contig bmc multi LaTeX xpression ssembly rees Rao
% LocalWords:  Zhang Moler Dhananjai Mufit de novo introns NGS bruijn ABIITY bp
% LocalWords:  GenBank indels contigs TGICL Metasim MetaSim Runtime runtime pre
% LocalWords:  zebrafish runtimes ESTs paralilization
